# Lab 8. Points to Predictions

## Overview

## Research Question

## Environment Setup

```{r}
#setwd("~/Desktop/Lab8-CodeData")
```
```{r message=FALSE, warning=FALSE}
library(sf)
library(tmap)
library(leaflet)
library(raster) # Needed for grid and kernel density surface
library(adehabitatHR) # Needed for kernel density surface
```

## Load and Join Data

Load non-spatial Data (csv) <br />  

```{r}
Census.Data <-read.csv("practicaldata.csv")
head(Census.Data)
```

Load spatial Data (shapefile) <br />  

```{r}
Output.Areas <- st_read("Camden_oa11.shp")
head(Output.Areas)
```

Join our census data to the shapefile <br />  

```{r}
OA.Census <- merge(Output.Areas, Census.Data, by.x="OA11CD", by.y="OA")
head(OA.Census)
```

Load the house prices csv file <br />  

```{r}
houses <- read.csv("CamdenHouseSales15.csv")
head(houses)
```

## Inspect and Prepare Data

We only need a few columns for this practical <br />  

```{r}
houses <- houses[,c(1,2,8,9)]
head(houses)
```

create a House.Points SpatialPointsDataFrame using coordinates in columns 3 and 4 <br />  

```{r}
House.Points <- st_as_sf(houses, coords = c("oseast1m","osnrth1m"), crs = 27700)
```

Plot to ensure the points were encoded correctly <br />  

```{r}
plot(House.Points)
```

Check coordinate reference system, transform if needed. <br />  
```{r}
st_crs(OA.Census) 
st_crs(House.Points)
```

## Point Data Method: Basic Visualizations

This plots a blank base map, we have set the transparency of the borders to 0.4 <br />  
```{r}
tm_shape(OA.Census) + tm_borders(alpha=.4) 
```

Creates a color-coded dot map <br />  
```{r}
tm_shape(OA.Census) + tm_borders(alpha=.4) +
  tm_shape(House.Points) + tm_dots(col = "Price", palette = "Reds", style = "quantile") 
```

Rescale the points <br />  
```{r}
tm_shape(OA.Census) + tm_borders(alpha=.4) +
  tm_shape(House.Points) + tm_dots(col = "Price", scale = 1.5, palette = "Reds", style = "quantile", title = "Price Paid (£)")  
```

Turn on interactive Leaflet map view <br />  
```{r}
tmap_mode("view")
```

View interactive Map; in Export, Save as Web Page <br />  
```{r}
tmap_mode("plot")
```

## Point Data Method: Graduate Symbol Visualizations

Creates a graduated symbol map <br />  
```{r}
tm_shape(OA.Census) + tm_fill("Unemployed", alpha=0.8, palette = "Greys", style = "quantile", title = "% Unemployed") + 
  tm_borders(alpha=.4) + 
  tm_shape(House.Points) + tm_bubbles(size = "Price", col = "Price", palette = "PuRd", style = "quantile", legend.size.show = FALSE, title.col = "Price Paid (£)", border.col = "black", border.lwd = 0.1, border.alpha = 0.1) +
  tm_layout(legend.text.size = 0.8, legend.title.size = 1.1, frame = FALSE)
```

## Point Data Method: Buffer Generation

Create 200m buffers for each house point <br />  
```{r}
house_buffers <- st_buffer(House.Points, 200)
```

Map in tmap <br />  
```{r}
tm_shape(OA.Census) + tm_borders() +
  tm_shape(house_buffers) + tm_borders(col = "blue") +
  tm_shape(House.Points) + tm_dots(col = "red") 
```

## Point Data Method: Buffer Count per Area

Count buffers within each area; generates a vector of totals <br />  
```{r}
count_buffers <- lengths(st_within(OA.Census, house_buffers))
head(count_buffers)
```

Stick buffer totals back to the census master file <br />  
```{r}
OA.Census <- cbind(OA.Census,count_buffers)
head(OA.Census)
```

Map density of buffers per census area <br />  
```{r}
tm_shape(OA.Census) + tm_fill(col = "count_buffers", palette = "BuGn", style = "quantile",title = "Housing Density")
```

## Point Data Method: Buffer Union (or Dissolve)
Merge the Buffers <br />  
```{r}
union.buffers <- st_union(house_buffers)
```
Map Housing Buffers <br />  
```{r}
tm_shape(OA.Census) + tm_borders() +
  tm_shape(union.buffers) + tm_fill(col = "blue", alpha = .4) + tm_borders(col = "blue") +
  tm_shape(House.Points) + tm_dots(col = "red") 
```

## Point Data Method: Group Attribute by Area
Point in polygon. Gives the points the attributes of the polygons that they are in <br />  
```{r}
pip <- st_join(House.Points, OA.Census, join = st_within)
head(pip)
```

Aggregate average house prices by the OA11CD (OA names) column <br />  
```{r}
OA <- aggregate(pip$Price, by = list(pip$OA11CD), mean)
head(OA)
```

Change the column names of the aggregated data <br />  
```{r}
names(OA) <- c("OA11CD", "Price")
```

Join the aggregated data back to the OA.Census polygon <br />  
```{r}
OA.Census <- merge(OA.Census, OA, by = "OA11CD", all.x = TRUE)
```

Map mean housing price per area <br />  
```{r}
tm_shape(OA.Census) + tm_fill(col = "Price", style = "quantile", title = "Mean House Price (£)")
```

## Point Data Method: Group by Grid
Convert sf objects to sp objects for raster and adehabitate packages <br />  
```{r}
OA.Census.sp <- sf:::as_Spatial(OA.Census)
House.Points.sp <- sf:::as_Spatial(House.Points)
```
Make a Grid. First define boundaries or extent of grid <br />  
```{r}
grid.extent <- extent(bbox(OA.Census.sp))        
```
Next grade a raster object from the extent <br />  
```{r}
grid.raster <- raster(grid.extent)            
```
Specify our dimensions; split extent into 30 x 30 cells <br />  
```{r}
dim(grid.raster) <- c(50,50)       
```
Project the grid using our shapefile CRS <br />  
```{r}
projection(grid.raster) <- CRS(proj4string(OA.Census.sp))  
```
Convert into a spatial data frame <br />  
```{r}
grid.sp <- as(grid.raster, 'SpatialPolygonsDataFrame') 
```
Convert spatial data frame to matrix data structure <br />  
```{r}
grid.matrix <- grid.sp[OA.Census.sp,]
```
Aggregate housing prices by grid matrix <br />  
```{r}
OA.grid <- aggregate(x=House.Points.sp["Price"], by=grid.matrix,FUN=mean)
```
If there is a null value, assign 0 <br />  
```{r}
OA.grid$Price[is.na(OA.grid$Price)] <- 0
```
Inspect data range <br />  
```{r}
summary(OA.grid@data)
```
Map <br />  
```{r}
tm_shape(OA.grid) + tm_fill(col = "Price", style = "quantile", n = 7, title = "Mean House Price (£)") 
```
## Point Data Method: Kernel Density Surface
Runs the kernel density estimation; look up the function parameters for more options
```{r}
kde.output <- kernelUD(House.Points.sp, h="href", grid = 1000)
plot(kde.output)
```

Converts to raster <br />  
```{r}
kde <- raster(kde.output)
```

Sets projection to British National Grid <br />  
```{r}
projection(kde) <- CRS("+init=EPSG:27700")
```

Maps the raster in tmap, "ud" is the density variable <br />  
```{r}
tm_shape(kde) + tm_raster("ud")
```

Creates a bounding box based on the extents of the Output.Areas polygon <br />  
```{r}
bounding_box <- bbox(OA.Census.sp)
```

Maps the raster within the bounding box <br />  
```{r}
tm_shape(kde, bbox = bounding_box) + tm_raster("ud")
```

Mask the raster by the output area polygon <br />  
```{r}
masked_kde <- mask(kde, Output.Areas)
```

Maps the masked raster, also maps white output area boundaries <br />  
```{r}
tm_shape(masked_kde, bbox = bounding_box) + tm_raster("ud", style = "quantile", n = 100, legend.show = FALSE, palette = "YlGnBu") +
  tm_shape(Output.Areas) + tm_borders(alpha=.3, col = "white") +
  tm_layout(frame = FALSE)
```

Compute homeranges for 75%, 50%, 25% of points, objects are returned as spatial polygon data frames <br />  
```{r}
range75 <- getverticeshr(kde.output, percent = 75)
range50 <- getverticeshr(kde.output, percent = 50)
range25 <- getverticeshr(kde.output, percent = 25)
```

The code below creates a map of several layers using tmap <br />  
```{r}
tm_shape(Output.Areas) + tm_fill(col = "#f0f0f0") + tm_borders(alpha=.8, col = "white") +
  tm_shape(House.Points) + tm_dots(col = "blue") +
  tm_shape(range75) + tm_borders(alpha=.7, col = "#fb6a4a", lwd = 2) + tm_fill(alpha=.1, col = "#fb6a4a") +
  tm_shape(range50) + tm_borders(alpha=.7, col = "#de2d26", lwd = 2) + tm_fill(alpha=.1, col = "#de2d26") +
  tm_shape(range25) + tm_borders(alpha=.7, col = "#a50f15", lwd = 2) + tm_fill(alpha=.1, col = "#a50f15") +
  tm_layout(frame = FALSE)
```

Write your kernel density surface to raster format <br />  
```{r}
writeRaster(masked_kde, filename = "kernel_density.grd")
```
